\subsection{Transfer Learning and Fine-Grained Classification}

Adapting a large pre-trained backbone to a small target dataset usually means full fine-tuning: take the ImageNet weights and update everything end-to-end on the new task. \citet{yosinski2014transferable} found that lower layers learn general features that transfer across datasets, while higher layers become more task-specific. The practical recipe follows from that observation: start from ImageNet, optionally freeze the early layers, and fine-tune the rest.

CNNs went through a fast evolution after AlexNet \citep{krizhevsky2012alexnet} won ImageNet in 2012. VGG \citep{simonyan2014vgg} stacked $3 \times 3$ convolutions to push depth, but plain stacking ran into vanishing gradients. Residual networks \citep{he2016resnet} fixed that by adding skip connections so gradients could bypass entire blocks, which let the network grow to 152 layers. Later families like DenseNet \citep{huang2017densenet} (every layer connected to every other) and EfficientNet \citep{tan2019efficientnet} (compound scaling of depth, width, and resolution) refined the design. We use ResNet18 as the CNN side of this report because it is the smallest member of the family that consistently transfers well, which keeps each training run cheap enough to sweep many configurations.

Vision Transformers \citep{dosovitskiy2021vit} arrived later and showed that pure attention can match or beat CNNs on image classification, provided the model is pre-trained on enough data. ViT splits an image into $16 \times 16$ patches, embeds each patch as a token, and runs them through a standard Transformer encoder with a special classification token at the front. Our VPT experiments build on a ViT-B/16 because it is the smallest variant the original VPT paper used and remains the standard reference backbone for prompt tuning work.

Fine-grained recognition is harder than generic object recognition because inter-class differences are subtle and intra-class variation is large. Bilinear CNNs \citep{lin2015bilinear} capture pairwise feature interactions and improved fine-grained accuracy by computing the outer product of two streams of CNN features. The Oxford Flowers 102 dataset itself was introduced by \citet{nilsback2008}, who reported 72.8\% accuracy with a multiple-kernel SVM combining SIFT, HOG, and HSV descriptors. The paper established mean per-class accuracy as the primary evaluation metric, which we adopt to handle the class-imbalanced test split.

\subsection{Advanced Convolution Operators}

Several generalisations of the standard $3 \times 3$ convolution have shown up in the literature, each motivated by a different limitation.

Dilated convolution was introduced by \citet{yu2016dilated} for dense prediction. It inserts zeros between kernel weights to enlarge the receptive field without adding parameters. \citet{chen2017deeplab} used dilated convolution in DeepLab to keep spatial resolution while still capturing wider context. Wider context is what we want from the deeper layers of ResNet18, so that is where we test the operator.

Deformable convolution \citep{dai2017deformable} replaces the rigid sampling grid with one that adapts per pixel. A small auxiliary branch predicts $2 \cdot k^2$ offsets per spatial location, and the main convolution samples at the shifted positions. The original paper showed gains on object detection and segmentation, where target objects deform significantly. Flowers vary a lot in petal arrangement and pose, so it seemed worth trying.

Depthwise-separable convolution factors a $3 \times 3$ convolution into a per-channel spatial filter followed by a $1 \times 1$ mixing. \citet{howard2017mobilenets} used the factorisation in MobileNet to cut compute by about $8\times$ for mobile deployment. \citet{chollet2017xception} pushed further with Xception and showed depthwise-separable can match or beat standard convolution on ImageNet given enough capacity. We tried it on Flowers 102 because a smaller parameter count seemed like a reasonable hedge against overfitting at 10 examples per class.

\subsection{Parameter-Efficient Tuning}

The idea of adapting a large pre-trained model with very few extra parameters started in NLP. GPT-3 \citep{brown2020gpt3} showed that a frozen language model could perform many downstream tasks with no gradient updates at all by just conditioning on a natural-language prompt. \citet{li2021prefix} and \citet{lester2021prompt} replaced the natural-language prompt with continuous learnable vectors prepended to the input, calling the technique prefix tuning and prompt tuning respectively, and showed it matched full fine-tuning on text classification with under 1\% of the trainable parameters. LoRA \citep{hu2022lora} took a different route, injecting low-rank updates into the attention weights, and is now widely used to adapt large language models on consumer hardware.

In vision, the analogous landscape is younger. Full fine-tuning still updates every backbone parameter and stores a separate copy of the model per downstream task. Lighter alternatives have shown up over the past few years. Linear probing freezes the backbone and trains only a linear classifier on top of its features. BitFit updates only the bias terms. Adapter layers \citep{jia2022vpt} insert small bottleneck modules between Transformer layers and train only those modules. Each method trades plasticity for parameter efficiency.

Visual Prompt Tuning (VPT) \citep{jia2022vpt} brings the prompt-tuning idea from language models to vision. Instead of modifying the network, VPT inserts learnable prompt vectors into the Transformer's input space. The backbone stays frozen and only the prompts plus the classifier head are trained. There are two variants: shallow injects prompts at the first Transformer layer only, deep injects fresh prompts at every layer. The original paper found that VPT-Deep beat full fine-tuning on 20 of 24 downstream tasks while training less than 1\% of the parameters, and the gap was largest in low-data regimes. That low-data behaviour is why we picked VPT as the main alternative to ResNet18 fine-tuning on Flowers 102.

A 2025 follow-up, Prompt-CAM \citep{chowdhury2025promptcam}, repurposed the prompts as an interpretability tool: each class-specific prompt's attention map highlights the discriminative regions of the input. We do not implement Prompt-CAM, but it is an obvious next step now that we already train VPT models.

\subsection{Self-Supervised Pre-Training}

The pre-trained ViT-B/16 we use is supervised on ImageNet-21k, but a parallel line of work removes the need for labels at pre-training time. Contrastive methods like MoCo and SimCLR pull augmented views of the same image together and push views of different images apart. \citet{caron2021dino} pushed this further with DINO, which trains a student-teacher pair on a ViT without labels and produces features whose attention maps already segment objects. Masked autoencoders (MAE) \citep{he2022mae} take a different angle: mask 75\% of the patches in an input image and train the model to reconstruct them, which yields features that transfer well to downstream classification. We do not use self-supervised backbones in this report (the supervised ViT-B/16 from \texttt{torchvision} was simpler to obtain), but the literature suggests they would likely help VPT more than they hurt, especially on fine-grained datasets where labelled pre-training data is scarce.

\subsection{Few-Shot Learning}

Few-shot learning is the problem of generalising from very few labelled examples per class. The literature splits roughly into metric learning, meta-learning, and plain transfer-learning baselines.

Metric learning embeds inputs into a feature space where same-class examples are close and different-class examples are far apart, then classifies new examples by nearest neighbour. \citet{vinyals2016matching} introduced Matching Networks, which compute attention-weighted nearest-neighbour predictions over a small support set. \citet{snell2017prototypical} simplified the idea with Prototypical Networks: each class is the centroid of its support embeddings, and queries are classified by Euclidean distance to those centroids. The triplet loss we use to fine-tune VPT-Deep is from the same family; it pulls same-class embeddings together and pushes different-class embeddings apart.

Meta-learning treats each few-shot task as a training instance and optimises for fast adaptation. MAML \citep{finn2017maml} learns an initialisation from which a few gradient steps on a new task produce a good model. These methods need episodic training across many tasks, which is not directly applicable to a single dataset like Flowers 102, so we do not implement them.

Transfer learning baselines just fine-tune a pre-trained backbone on the few-shot training set. Recent work has shown that strong pre-training plus standard fine-tuning is enough to match more elaborate few-shot methods on many datasets. Our results agree: VPT-Deep with no episodic training wins at every $k$ we tried.

\subsection{Regularization for Small Datasets}

Standard cross-entropy training with one-hot targets pushes the network toward overconfident predictions, which leads to overfitting when data is scarce. Label smoothing \citep{szegedy2016rethinking} replaces the one-hot target $y$ with $\tilde{y}_i = (1 - \alpha) y_i + \alpha / K$, where $K$ is the number of classes. This bounds the maximum softmax probability the model can predict and slightly improves calibration in addition to acting as a regularizer.

MixUp \citep{zhang2018mixup} trains on convex combinations of pairs of inputs and labels: $\tilde{x} = \lambda x_i + (1-\lambda) x_j$, $\tilde{y} = \lambda y_i + (1-\lambda) y_j$, with $\lambda \sim \text{Beta}(\alpha, \alpha)$. Training on these mixed examples enforces locally linear behaviour between training points and reduces memorisation. CutMix \citep{yun2019cutmix} is a follow-up that cuts a rectangular patch from one image and pastes it onto another, with the label mixed proportionally to the patch area. We use the official \texttt{torchvision} v2 implementation of MixUp.

Triplet loss \citep{schroff2015facenet} trains an embedding by pushing the distance from an anchor to a positive (same class) sample below its distance to a negative (different class) sample by some margin. \citet{hermans2017triplet} proposed batch-hard mining, which picks the hardest positive and hardest negative within each batch and gives much stronger training signal than random or semi-hard sampling. Our triplet experiments use batch-hard mining.
